Linear stability of magnetized massive protoplanetary disks

### 2014-05-28 paper/cover.tex

I am writing to submit the second revised manuscript for `Non-barotropic linear Rossby wave instability
in three-dimensional disks'. I include a list of revisions made in light of the reviewer's comments.

>>>>

I am writing to submit the manuscript `Linear stability of magnetized massive protoplanetary disks'.

### 2014-05-28 paper/results.tex
minor edits

\subsubsection{Uniform
  resistivity}   
Fig. \ref{compare_growth_poly_uniresis} plots MRI growth rates as a
function of $Q$ and $\Lambda_0$. The resistivity is uniform ($A=1$). 
For ideal MHD ($\Lambda_0>1$), there is negligible dependence on
$Q$. In the resistive limit   
($\Lambda_0<1$), however, growth rates decrease noticeably for $Q<0.5$
($Q_\mathrm{2D}\lesssim 1.5$).  Since we find density and potential
perturbations to be negligible (i.e. the linear response is
non-self-gravitating), this shows that disk self-gravity can affect the
MRI through the background equilibrium.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth_poly_uniresis2}
  \caption{MRI growth rates as a function of $Q$ and midplane 
    Elsasser numbers, in polytropic disks with $\beta=100$ and uniform
    resistivity.  
    \label{compare_growth_poly_uniresis}}
\end{figure}

@@

In the ideal limit, we find $\lambda < 2H$ throughout most of the disk
for the values of $Q$ considered, so self-gravity does not affect
growth rates significantly. However, the ratio $\lambda/2H$ does
increase with stronger self-gravity. Consequently, the wavelength of
the instability, in units of $H$, increases. This is shown in
Fig. \ref{compare_result_lambda10} which plots the magnetic energies
for $\Lambda_0=10$ and a range of $Q$ values. The number of vertical
nodes decrease with $Q$, i.e. the disk accommodates fewer wavelengths
because increasing vertical self-gravity makes it thinner.

@@

Self-gravity appreciably decreases the MRI growth rates in the
resistive limit. Fig. \ref{lambda_poly_resis} plots 
Eq. \ref{sano_crit} for $\Lambda_0=0.3$. In the non-self-gravitating
disk ($Q=4$) the instability criterion is marginally satisfied and the
MRI operates. As $Q$ decreases, 
Eq. \ref{sano_crit} is violated and the MRI growth rate is
significantly reduced. This is seen for $Q=0.2$ where $\lambda \geq 2H$ throughout
the disk. (The instability is not suppressed since
Eq. \ref{lambda_ideal}---\ref{lambda_resis} is only exact for
unstratified disks.) Although the function 
$f(Q)$ does not change significantly for the range of $Q$ considered,
the dependence of $\lambda_\mathrm{resis}$ on $f(Q)$ is amplified by the
denominator $\Lambda_0<1$ in the resistive case. Modes in
Fig. \ref{lambda_poly_resis} have no nodes in the magnetic energy
$E_m$ except at $z\simeq0,\,H$, i.e. only the longest wavelength survives
with large resistivity. 
%self-gravity tips it over

@@

@@

\begin{align}
  &  0= v_A^2k^2\dvx + \imgi\sigma\left(\imgi\sigma \dvx - 2\Omega\dvy + \imgi k_x \w\right),\label{simp1}\\
  &  0= v_A^2k_z^2\left(\dvy + \frac{\imgi S
  }{\sigma}\dvx\right) + \imgi\sigma\left(\imgi\sigma\dvy +
  \frac{\kappa^2}{2\Omega}\dvx\right),\label{simp2}\\
  & 0 = -k_z^2\w + \frac{\sigma^2}{c_s^2}W + \sigma k_x \dvx, \label{simp3}\\
  & 0 = k^2\dphi + \frac{\Omega^2}{c_s^2Q}W,\label{simp4}
\end{align}
where $k^2 = k_z^2 + k_x^2$. We imagine an iterative procedure to
solve the above equations, starting from the Cowling approximation
where $\delta\Phi=0$. This is the standard MRI and we denote the
solution as $\dvx^{(0)}$, $\dvy^{(0)}$ and $W^{(0)}$. Eq. \ref{simp3} implies
\begin{align}
  W^{(0)} = \frac{c_s^2\sigma k_x \dvx^{(0)}}{c_s^2k_z^2 - \sigma^2}. 
\end{align}
We argue below that $c_s^2k_z^2\gg \sigma^2$ by taking $k_z\sim
\Omega/v_A$. Then, recalling $W=c_s^2\delta\rho/\rho$, we can write
\begin{align}
  \frac{\delta\rho^{(0)}}{\rho} \sim
  \frac{\sigma}{\Omega}\frac{1}{\beta}\left[\frac{k_x\dvx^{(0)}}{\Omega}\right].   
\end{align}
The MRI has, in general, a non-zero density perturbation. However, it
is negligible for $k_x\to 0$ and/or a weak field ($\beta \gg 1$).

>>>>

\subsubsection{Uniform
  resistivity}   
Fig. \ref{compare_growth_poly_uniresis} plots MRI growth rates as a
function of $Q$ and $\Lambda_0$. The resistivity is uniform ($A=1$). 
For ideal MHD and {\bf a weak field ($\Lambda_0>1$, $\beta=100$)}, there is
negligible dependence on $Q$. {\bf However, with $\beta=25$ or 
in the resistive limit  ($\Lambda_0<1$)}, growth rates decrease noticeably for $Q<0.5$
($Q_\mathrm{2D}\lesssim 1.5$).  Since we find density and potential
perturbations to be negligible (i.e. the linear response is
non-self-gravitating), this shows that disk self-gravity can affect the
MRI through the background equilibrium.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth_poly_uniresis2}
  \caption{{\bf MRI growth rates as a function of $Q$ and midplane 
    Elsasser numbers $\Lambda_0$, in polytropic disks with $\beta=100$
    (solid, dotted and dot-dashed lines). The dash-triple dot line is
    the $\Lambda_0=10$ case with $\beta=25$. The resistivity is
    uniform.}   
    \label{compare_growth_poly_uniresis}}
\end{figure}

@@

In the ideal limit {\bf with $\beta=100$}, we find $\lambda < 2H$
throughout most of the disk for the values of $Q$ considered, so
self-gravity does not affect growth rates significantly. However, the
ratio $\lambda/2H$ does increase with stronger
self-gravity. Consequently, the wavelength of the instability, in
units of $H$, increases. This is shown in
Fig. \ref{compare_result_lambda10} which plots the magnetic energies
for $\Lambda_0=10$ and a range of $Q$ values. The number of vertical 
nodes decrease with $Q$, i.e. the disk accommodates fewer wavelengths
because increasing vertical self-gravity makes it thinner.

{\bf
We repeated the $\Lambda_0=10$ case with a stronger field $\beta=25$,
shown in Fig. \ref{compare_growth_poly_uniresis} as the dashed-triple
dot line. Here, strong self-gravity is effective in reducing the
growth rate, because decreasing $\beta$ enhances the dependence of
$\lambda/H$ on $f(Q)$. 
}

@@

Self-gravity {\bf also} appreciably decreases the MRI growth rates in the
resistive limit. Fig. \ref{lambda_poly_resis} plots 
Eq. \ref{sano_crit} for $\Lambda_0=0.3$. In the non-self-gravitating
disk ($Q=4$) the instability criterion is marginally satisfied and the
MRI operates. As $Q$ decreases, 
Eq. \ref{sano_crit} is violated and the MRI growth rate is
significantly reduced. This is seen for $Q=0.2$ where $\lambda \geq 2H$ throughout
the disk. (The instability is not suppressed since
Eq. \ref{lambda_ideal}---\ref{lambda_resis} is only exact for
unstratified disks.) Although the function 
$f(Q)$ does not change significantly for the range of $Q$ considered,
the dependence of {\bf $\lambda/H$} on $f(Q)$ is amplified by the
denominator $\Lambda_0<1$ in the resistive case. Modes in
Fig. \ref{lambda_poly_resis} have no nodes in the magnetic energy
$E_m$ except at $z\simeq0,\,H$, i.e. only the longest wavelength survives
{\bf against} large resistivity. 
%self-gravity tips it over

@@

{\bf 
%We find qualitatively similar results for large $A$ with small
%  $\Lambda_0$. 
We also performed additional calculations with $A=10^3$ and $A=10^4$
\citep[see Fig. 1 of ][]{gressel12}. For $\Lambda_0=0.1$, we find no
significant increase in the magnetic energy in the resistive zones. 
However, lowering $\Lambda_0$ gives similar
results to Fig. \ref{poly_layer}, e.g. for $A=10^3$ and
$\Lambda_0=10^{-2}$ or $A=10^4$ and $\Lambda_0=10^{-3}$ the magnetic
energy penetrates into the resistive zone for strongly
self-gravitating disks. In general, the magnetic energy density maximum moves
toward the midplane with increasing self-gravity.    
}

@@

\begin{align}
  &  0= v_A^2k^2\dvx + \imgi\sigma\left(\imgi\sigma \dvx - 2\Omega\dvy + \imgi k_x \w\right),\label{simp1}\\
  &  0= v_A^2k_z^2\left(\dvy + \frac{\imgi S
  }{\sigma}\dvx\right) + \imgi\sigma\left(\imgi\sigma\dvy +
  \frac{\kappa^2}{2\Omega}\dvx\right),\label{simp2}\\
  & 0 = -k_z^2\w + \frac{\sigma^2}{c_s^2}W + \sigma k_x \dvx, \label{simp3}\\
  & 0 = k^2\dphi + \frac{\Omega^2}{c_s^2Q}W,\label{simp4}
\end{align}
where $k^2 = k_z^2 + k_x^2$. We imagine an iterative procedure to
solve the above equations, starting from the Cowling approximation
where $\delta\Phi\to0$ and $Q\to\infty$. This is the standard MRI and we denote the
solution as $\dvx^{(0)}$, $\dvy^{(0)}$ and $W^{(0)}$. Eq. \ref{simp3} implies
\begin{align}
  W^{(0)} = \frac{c_s^2\sigma k_x \dvx^{(0)}}{c_s^2k_z^2 - \sigma^2}. 
\end{align}
We argue below that $c_s^2k_z^2\gg \sigma^2$ by taking $k_z\sim
\Omega/v_A$. Then, recalling $W=c_s^2\delta\rho/\rho$, we can write
\begin{align}
  \frac{\delta\rho^{(0)}}{\rho} \sim
  \frac{\sigma}{\Omega}\frac{1}{\beta}\left[\frac{k_x\dvx^{(0)}}{\Omega}\right].   
\end{align}
The MRI has, in general, a non-zero density perturbation. However, it
is negligible for $k_x\to 0$ and/or a weak field ($\beta \gg 1$).

### 2014-05-28 paper/setup.tex

\subsection{Resistivity profile}\label{resis_profile}
We adopt constant resistivity or a 
resistivity prescription such that $\eta(z)$ increases away from the
midplane. In the latter case, we follow \cite{fleming03} and use 
resistivity profile 
\begin{align}
  \eta(z) =
  \sqrt{2}\eta_0\left[\exp{\left(-g_+\right)}+\exp{\left(-g_-\right)}\right]^{-1/2},  
\end{align}
where
\begin{align}
  &g_\pm(z) =  \frac{\Sigma_\pm(z)-\Sigma_0}{\Sigma_*}, \\
  &\Sigma_\pm(z) = \int_{\pm z}^\infty\rho(z^\prime)dz^\prime, \label{sigma_pm}
\end{align}
and $\Sigma_0\equiv\Sigma_{\pm}(0)$, so that $g_\pm(0)=0$ and $\eta_0 
= \eta(0)$. The constant $\Sigma_*$ is chosen such that 
\begin{align}
  \cosh{\left(\frac{\Sigma_0}{\Sigma_*}\right)} =
  \left[\frac{\eta_0}{\eta(\infty)}\right]^2,
\end{align}
and we define $\eta_0/\eta(\infty)\equiv A$ as the conductivity 
boost factor from the midplane to the disk surface. We remark that
once $\rho$ and $d\rho/dz$ are obtained from
Eq. \ref{eqm_eqns1} --- \ref{eqm_eqns2}, the integration for
Eq. \ref{sigma_pm} can be performed implicitly by using Poisson's 
equation.

>>>>

\subsection{Resistivity profile}\label{resis_profile}
We adopt constant resistivity or a 
resistivity prescription such that $\eta(z)$ increase {\bf towards} the
midplane. In the latter case, we follow \cite{fleming03} and use 
resistivity profile 
\begin{align}
  \eta(z) =
  \sqrt{2}\eta_0\left[\exp{\left(-g_+\right)}+\exp{\left(-g_-\right)}\right]^{-1/2},  
\end{align}
where
\begin{align}
  &g_\pm(z) =  \frac{\Sigma_\pm(z)-\Sigma_0}{\Sigma_*}, \\
  &\Sigma_\pm(z) = \int_{\pm z}^\infty\rho(z^\prime)dz^\prime, \label{sigma_pm}
\end{align}
and $\Sigma_0\equiv\Sigma_{\pm}(0)$, so that $g_\pm(0)=0$ and $\eta_0 
= \eta(0)$. The constant $\Sigma_*$ is chosen such that 
\begin{align}
  \cosh{\left(\frac{\Sigma_0}{\Sigma_*}\right)} =
  \left[\frac{\eta_0}{\eta(\infty)}\right]^2,
\end{align}
and we define $\eta_0/\eta(\infty)\equiv A$ as the conductivity 
boost factor from the midplane to the disk surface. We remark that
once $\rho$ and $d\rho/dz$ are obtained from
Eq. \ref{eqm_eqns1} --- \ref{eqm_eqns2}, the integration for
Eq. \ref{sigma_pm} can be performed implicitly by using Poisson's 
equation.
